### 2024-04-11 docs/SOSP 2024/reduction.tex
Edit reduction section

\section{Reductions : Representation, Optimization and Lowering}

Currently, existing reduction support in MLIR is insufficient 
to code generate and optimize the reductions \Treebeard{} needs
to perform while performing inference (sum up individual 
tree predictions to compute the prediction of the model). 
MLIR only supports reductions of value types and cannot 
directly represent and optimize inplace reductions of 
several elements of a memory buffer.

We design a mechanism to specify accumulating values into
an element of a multi-dimensional array inplace. 
The accumulation is performed inside an arbitrary 
loop nest where several surrounding loops maybe parallel and 
the ultimate target machine maybe a CPU or a GPU. Because this problem 
is of general interest, we design this as an MLIR dialect.

The main abstraction we introduce is the \op{reduce} op. It models 
atomically accumulating values into an element of a 
multi-dimensional array (represented by a MLIR \op{memref}).
The following example sums up the elements of the 1D memref \op{arr}
into the first element of the memref \op{result}. It does this
using in two concurrent iterations of a surrounding 
parallel loop.

\begin{lstlisting}[style=c++]
  builtin.func @ReduceVector(%arr: memref<num_elemsxf64>, %result: memref<1xf64>) -> void {
    par.for i0 in range(0 : num_elems/2 : num_elems) {
      for i1 in range(0 : num_elems/2) 
        reduce(%result, 0, arr[i0 + i1]) <"+", 0.0>
    }
  }
\end{lstlisting}

The semantics of the \op{reduce} op guarantee that all elements are 
correctly added and that there is no race between the parallel 
iterations of the loop.

The \op{reduce} op is defined for all associative and commutative
reduction operations with a well-defined initial value. The 
reduction operator and the initial value are attributes applied
on the \op{reduce} op.

The main differences between our \op{reduce} and the existing 
reductions in MLIR are the following:
1. Existing reductions only support scalar, by value reductions. 
It does not support accumulating inplace into memref elements.
2. Existing reduction support in MLIR does not provide a unified
way to handle reductions in GPUs and CPUs. To the best of our 
knowledge, there is currently no way to model reductions on GPUs
in MLIR.
2. Existing reductions have strict rules about where they can be 
written. For example, \op{scf.reduce} needs to be an immediate 
child \op{scf.parallel}. However, our \op{reduce} op can be
generated anywhere in the loop nest.

Having modeled the reductions with an abstract op, the 
aim now is to lower this to a correct and optimized 
implementation on both CPU and GPU. In order to do this,
we make the following observations. 
1. The \op{reduce} op has a loop carried dependency on itself
and loop carried dependences on other \op{reduce} ops that
accumulate into the same target array. 
A simple lowering to a sequence of load-add-store instructions is 
incorrect if any of these dependences are carried by a parallel 
loop. We call any such surrounding parallel loop a 
\textbf{conflicting loop} (\TODO{Change the name!}) for the reduction.
2. There is a race between the parallel iterations of such a loop 
when naively accumulating values into target memref elements. 
To avoid this race, the result memref can be \textbf{privatized} wrt each 
surrounding conflicting loop. Subsequently, each privatized dimension 
can be reduced at the end of the conflicting loop it was inserted for. 
\TODO{We cannot do better than this in terms of memory usage} \TODO{Need a proof}.

\begin{definition}
A parallel loop surrounding one or more \op{reduce} ops is 
a \textbf{conflicting loop} for a target multi-dimensional array if this 
loop has a non-zero dependence distance for the dependence between
any of the contained \op{reduce} ops.
\end{definition}

In the context of \Treebeard{}, this set of loops is exactly the set 
of surrounding parallel loops that are iterating over trees. The 
results can be privatized for each conflicting loop iteratively 
and reductions along each privatized dimension can be inserted  
immediately following the loop the dimension was inserted due to.

We illustrate this process through the example above. The \op{i0} loop is 
a conflicting loop for the reduction into the \op{result} array.
We would therefore privatize the \op{result} memref for each iteration of 
the \op{i0} loop. 
\begin{lstlisting}[style=c++]
  builtin.func @ReduceVector(%arr: memref<num_elemsxf64>, %result: memref<1xf64>) -> void {
    results_1 = memref<2x1xf64>
    par.for i0 = range(0 : num_elems/2 : num_elems) {
      for i1 = range(0 : num_elems/2) 
        reduce(%result_1[i0/(num_elems/2), 0], arr[i0 + i1]) <"+", 0.0>
    }
    results = reduce_dimension(results_1, 0) <"+", 0.0>
  }
\end{lstlisting}
The op \op{reduce\_dimension} reduces values across the specified
dimension of an n-dimensional memref. In the above example, 
the \op{reduce\_dimension} op is reducing across all elements 
of the first dimension (index 0). Therefore, in this case, it 
produces a result memref with a single element (the first dimension
with size 2 is collapsed).

\begin{definition}
  \op{\textbf{reduce\_dimension(targetMemref, memref, dim, [indices], [rangeStart], [rangeEnd])}}:
   Computes the reduction over the dimension specified by \op{dimension} and stores the 
   result in \op{targetMemref}. \op{[indices]} must be a vector of \op{dim} elements
   (or empty if the dimension being reduced is the first dimension). \op{[rangeStart]} 
   and \op{[rangeEnd]} represent the range of indices following the reduction dimension and 
   must have the same number of elements. If both are \op{null} (not passed), 
   all elements of these dimensions are reduced. The computation performed by the op is as follows.

  $targetMemref[\vec{\boldsymbol{indices}}, \vec{\boldsymbol{k}}] = \sum_{i=0}^{shape[dim]} memref[\vec{\boldsymbol{indices}}, i, \vec{\boldsymbol{k}}]\quad   \forall \vec{\boldsymbol{k}} \in \left[[rangeStart_0, rangeEnd_0), ... , [rangeStart_n, rangeEnd_n)\right]$
\end{definition}

Consider the following code with nested parallel loops. 
(A situation where trees are split across both
threads and thread blocks could result in such generated 
code in \Treebeard{}.)
\begin{lstlisting}[style=c++]
  builtin.func @ReduceVector(%arr: memref<num_elemsxf64>, %result: memref<1xf64>) -> void {
    par.for i0 = range(0 : num_elems/2 : num_elems) {
      par.for i1 = range(0 : num_elems/4 : num_elems/2) {
        for i2 = range(0 : num_elems/4) 
          reduce(%result, 0, arr[i0 + i1 + i2]) <"+", 0.0>
      }
    }
  }
\end{lstlisting}

Here, the \op{i0} and \op{i1} loops are conflicting loops wrt 
the \op{result} memref. We \textbf{legalize} the reduction 
by privatizing the \op{result} array wrt the \op{i0} and \op{i1} loops.
However, there are now two privatized dimensions and therefore, 
two dimensions need to be reduced to compute the final result.
This multi-stage reduction is what enables us to model 
hierarchical reductions.

The following code shows how the reduction above is legalized. 
We introduce a new op, \op{reduce\_dimension\_inplace} which 
reduces a dimension of the input memref and stores results 
in the same array. This helps saves memory by removing the 
need to create multiple intermediate arrays to store results.
Only the final dimension reduction uses the \op{reduce\_dimension} op.

\begin{lstlisting}[style=c++]
  builtin.func @ReduceVector(%arr: memref<num_elemsxf64>, %result: memref<1xf64>) -> void {
    results_1 = memref<2x2x1xf64>
    par.for i0 = range(0 : num_elems/2 : num_elems) {
      par.for i1 = range(0 : num_elems/4 : num_elems/2) {
        for i2 = range(0 : num_elems/4) 
          index0 = i0/(num_elems/2)
          index1 = i1/(num_elems/4)
          reduce(%result_1[index0, index1, 0], arr[i0 + i1 + i2]) <"+", 0.0>
      }
      // result_1[i0/(num_elems/2), 0] = sum(result_1[i0/(num_elems/2), :])
      reduce_dimension_inplace(%result_1, 1, i0/(num_elems/2)) 
    }
    // result = sum(result[:, 0])
    %result = reduce_dimension(%result_1, 0)
  }
\end{lstlisting}

@@

>>>>

\section{Reductions : Representation, Optimization and Lowering}
\label{sec:reduction}
\Treebeard{} needs to sum up individual tree predictions to compute 
the prediction of the model while performing inference. However,
generating fused reductions within arbitrary loop nests specified 
using \Treebeard{}'s scheduling language is non-trivial. We found 
that existing reduction support in MLIR is insufficient to code
generate and optimize these reductions. MLIR only supports reductions
of value types and does not provide ways to lower reductions to GPUs. 
To address this gap, we design an MLIR dialect that allows us to
specify accumulating values into an element of a multi-dimensional array
and can be lowered to CPU or GPU.

The main abstraction we introduce is the \op{reduce} op. It models 
atomically accumulating values into an element of a 
multi-dimensional array (represented by an MLIR \op{memref}).
The following example shows how the \op{reduce} op can be used to 
sum up the elements of an array in parallel.

\begin{lstlisting}[style=c++]
  float arr[10], result[1]
  par.for i0 = 0  to 10 step 5 {
    for i1 = 0 to 5 
      reduce(result[0], arr[i0 + i1]) <"+", 0.0>
  }
\end{lstlisting}

The semantics of the \op{reduce} op is exactly the semantics of 
an atomic accumulation, i.e. it guarantees that all accumulations 
are correctly performed even in the presence of parallel loops. 
The \op{reduce} op is defined for all associative and commutative
reduction operations with a well-defined initial value. The 
reduction operator and the initial value are attributes applied
on the \op{reduce} op.

Having modeled the reductions with an abstract operation, the 
aim now is to lower this to a correct and optimized 
implementation on both CPU and GPU. In order to do this, we 
first determine if any parallel loop iterations can accumulate 
into the same array element. We call such loops 
\emph{\textbf{reduction loops}}. If such loops exist, we 
\textbf{\emph{privatize}} the array for each iteration of the
loop. We call this process \textbf{\emph{legalization}}.
Subsequently, each privatized dimension 
can be reduced at the end of the reduction loop it was inserted for. 
\TODO{We cannot do better than this in terms of memory usage} \TODO{Need a proof}.

In our example above, parallel iterations of the \op{i0} loop accumulate 
into the same element of the \op{result} array.
We would therefore privatize the \op{result} array for each iteration of 
the \op{i0} loop as follows. 
\begin{lstlisting}[style=c++]
  float arr[10], result[1]
  float resultPriv[2][1]
  par.for i0 = 0  to 10 step 5 {
    for i1 = 0 to 5 
      reduce(resultPriv[i0/5][0], arr[i0 + i1]) <"+", 0.0>
  }
  result = reduce_dimension(resultPriv, 0)
\end{lstlisting}

The op \op{reduce\_dimension} reduces values across the specified
dimension of an n-dimensional array. In the above example, 
the \op{reduce\_dimension} op is reducing across all elements 
of the first dimension (dimension 0). Therefore, in this case, it 
produces a result memref with a single element (the first dimension
with size 2 is collapsed).

% \begin{definition}
%   A parallel loop surrounding one or more \op{reduce} ops is 
%   a \textbf{conflicting loop} for a target multi-dimensional array if this 
%   loop has a non-zero dependence distance for the dependence between
%   any of the contained \op{reduce} ops.
% \end{definition}

% \begin{definition}
%   \op{\textbf{reduce\_dimension(targetMemref, memref, dim, [indices], [rangeStart], [rangeEnd])}}:
%    Computes the reduction over the dimension specified by \op{dimension} and stores the 
%    result in \op{targetMemref}. \op{[indices]} must be a vector of \op{dim} elements
%    (or empty if the dimension being reduced is the first dimension). \op{[rangeStart]} 
%    and \op{[rangeEnd]} represent the range of indices following the reduction dimension and 
%    must have the same number of elements. If both are \op{null} (not passed), 
%    all elements of these dimensions are reduced. The computation performed by the op is as follows.

%   $targetMemref[\vec{\boldsymbol{indices}}, \vec{\boldsymbol{k}}] = \sum_{i=0}^{shape[dim]} memref[\vec{\boldsymbol{indices}}, i, \vec{\boldsymbol{k}}]\quad   \forall \vec{\boldsymbol{k}} \in \left[[rangeStart_0, rangeEnd_0), ... , [rangeStart_n, rangeEnd_n)\right]$
% \end{definition}

% Consider the following code with nested parallel loops. 
% (A situation where trees are split across both
% threads and thread blocks could result in such generated 
% code in \Treebeard{}.)
% \begin{lstlisting}[style=c++]
%   builtin.func @ReduceVector(%arr: memref<num_elemsxf64>, %result: memref<1xf64>) -> void {
%     par.for i0 = range(0 : num_elems/2 : num_elems) {
%       par.for i1 = range(0 : num_elems/4 : num_elems/2) {
%         for i2 = range(0 : num_elems/4) 
%           reduce(%result, 0, arr[i0 + i1 + i2]) <"+", 0.0>
%       }
%     }
%   }
% \end{lstlisting}

% Here, the \op{i0} and \op{i1} loops are conflicting loops wrt 
% the \op{result} memref. We \textbf{legalize} the reduction 
% by privatizing the \op{result} array wrt the \op{i0} and \op{i1} loops.
% However, there are now two privatized dimensions and therefore, 
% two dimensions need to be reduced to compute the final result.
% This multi-stage reduction is what enables us to model 
% hierarchical reductions.

% The following code shows how the reduction above is legalized. 
% We introduce a new op, \op{reduce\_dimension\_inplace} which 
% reduces a dimension of the input memref and stores results 
% in the same array. This helps saves memory by removing the 
% need to create multiple intermediate arrays to store results.
% Only the final dimension reduction uses the \op{reduce\_dimension} op.

% \begin{lstlisting}[style=c++]
%   builtin.func @ReduceVector(%arr: memref<num_elemsxf64>, %result: memref<1xf64>) -> void {
%     results_1 = memref<2x2x1xf64>
%     par.for i0 = range(0 : num_elems/2 : num_elems) {
%       par.for i1 = range(0 : num_elems/4 : num_elems/2) {
%         for i2 = range(0 : num_elems/4) 
%           index0 = i0/(num_elems/2)
%           index1 = i1/(num_elems/4)
%           reduce(%result_1[index0, index1, 0], arr[i0 + i1 + i2]) <"+", 0.0>
%       }
%       // result_1[i0/(num_elems/2), 0] = sum(result_1[i0/(num_elems/2), :])
%       reduce_dimension_inplace(%result_1, 1, i0/(num_elems/2)) 
%     }
%     // result = sum(result[:, 0])
%     %result = reduce_dimension(%result_1, 0)
%   }
% \end{lstlisting}

@@

In the context of \Treebeard{}, the set of reduction loops is exactly the set 
of surrounding parallel loops that are iterating over trees. The 
results can be privatized for each loop iteratively 
and reductions along each privatized dimension can be inserted  
immediately following the loop the dimension was inserted due to.
